---
title: "DRD-RuralUrban"
author: 'Anisha Bhattarai'
date: "2026-07-04"
output: html_document
---
<!-- This script of DRD data track each LSOA appeared in top 50 most deprived areas across six WIMD publication since 2005, invisible communities chart become central to policy recommendation-->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(janitor)
```

## Step 1- Load master dataset
```{r load_data}
master <- read_csv("../data/processed/master_dataset.csv") %>%
  mutate(
    rural_urban = factor(rural_urban, levels = c("Rural", "Urban")),

    drd_category = factor(drd_category,
      levels = c(
        "LSOAs never ranked in top 50 most deprived",
        "LSOAs sometimes in the top 50 most deprived",
        "LSOAs often in the top 50 most deprived",
        "LSOAs in Deep-rooted deprivation"
      ),
      ordered = TRUE)
  )
# threshold from N stay correct dataset ever size change
N   <- nrow(master)
T10 <- floor(N * 0.10)   # = 191
T20 <- floor(N * 0.20)   # = 383
cat("N =", N, "| worst 10% <= ", T10, "| worst 20% <=", T20, "\n")


```

## Step 2- rural urban in each DRD categories
```{r drd_crosstab}
# Number of rural, urban LSOAs in each DRD category
table(master$drd_category, master$rural_urban)
```

## Step 3- % share within rural and urban
```{r drd_pct}
# row percentage-share of each category in rural and urban
prop.table(
  table(master$drd_category, master$rural_urban),
  margin = 1
) %>%
  round(3) * 100

```

## Step 4- rural in deep rooted  
```{r rural_deep_rooted}

master %>%
  filter(drd_category == "LSOAs in Deep-rooted deprivation",
         rural_urban  == "Rural") %>%
  select(lsoa_name, la_name, ruc_name,
         wimd_2025_rank, income_rank, access_rank, housing_rank)


```

## Step 5- invisible communities
```{r invisible_lsoas}
# Worst 10% for access in wales and never enter DRD tracking 
master %>%
  filter(rural_urban == "Rural",
         access_rank <= T10,    # worst 10% on access
         wimd_2025_rank > 50) %>%  # but not overall top 50
  summarise(
    count               = n(),
    median_overall_rank = median(wimd_2025_rank, na.rm=TRUE),
    median_access_rank  = median(access_rank,    na.rm=TRUE),
    median_income_rank  = median(income_rank,    na.rm=TRUE)
  )


```


## Stage 6- Visibility mechanism 
```{r mechanism_proof}
# structural proof- access deprivation alone vs income deprivation 
master %>%
  filter(rural_urban == "Rural") %>%
  mutate(dep_type = case_when(
    access_rank <= T20 & income_rank <= T20 ~ "Both access AND income",
    access_rank <= T20 & income_rank >  T20 ~ "Access deprived only",
    access_rank >  T20 & income_rank <= T20 ~ "Income deprived only",
    TRUE                                    ~ "Neither"
  )) %>%
  group_by(dep_type) %>%
  summarise(
    n                   = n(),
    median_overall_rank = median(wimd_2025_rank, na.rm=TRUE),
    pct_worst20_overall = round(100 * mean(wimd_2025_rank <= T20), 1),
    .groups = "drop"
  ) %>%
  arrange(median_overall_rank)


```

## Stage 7- Stability across 20 years
```{r drd_volatility}
# rank stability across 20 years by DRD category
# SD across 6 iterations
master %>%
  rowwise() %>%
  mutate(rank_sd = sd(c_across(c(wimd_2025_rank, wimd_2019_rank, wimd_2014_rank,
                                  wimd_2011_rank, wimd_2008_rank, wimd_2005_rank)),
                      na.rm=TRUE)) %>%
  ungroup() %>%
  group_by(drd_category) %>%
  summarise(n = n(), median_sd = round(median(rank_sd, na.rm=TRUE), 0), .groups="drop")

```

## Step 8- Persistent deprivation missing rural areas
```{r drd_domain_profile}
# compare income and access rank "never tracked"rural against "deep rooted" urban areas
master %>%
group_by(drd_category, rural_urban) %>%
  summarise(
    n             = n(),
    median_income = median(income_rank,  na.rm=TRUE),
    median_access = median(access_rank,  na.rm=TRUE),
    .groups = "drop"
  ) %>%
  filter(n >= 2) %>%
  arrange(drd_category, rural_urban)

```

## Step 9- Top 50 LSOA name and rural vs urban  
```{r top_50_deprived areas}
top50 <- master %>% slice_min(wimd_2025_rank, n = 50)
top50 %>%
  select(wimd_2025_rank, lsoa_name, la_name,
         rural_urban, income_rank, employment_rank, access_rank) %>%
  arrange(wimd_2025_rank)
```
 
## Step 10- Domain wise top 50 median rank
```{r domain_top50_domainrank}
dom_cols <- c("income_rank","employment_rank","health_rank",
              "education_rank","access_rank","housing_rank")

list(
  "Top 50 Rural (n=2)"      = top50 %>% filter(rural_urban == "Rural"),
  "Top 50 Urban (n=48)"     = top50 %>% filter(rural_urban == "Urban"),
  "All Rural Wales (n=561)" = master %>% filter(rural_urban == "Rural")
) %>%
  map_dfr(~summarise(.x, across(all_of(dom_cols),
                                ~round(median(.x, na.rm=TRUE), 0))),
          .id = "group")
```

## Step 11-Top 50 domain distribution of rural vs urban 
```{r domain_distribute_top50}
domain_cols   <- c("wimd_2025_rank","income_rank","employment_rank",
                   "health_rank","education_rank","access_rank",
                   "housing_rank","community_safety_rank","physical_env_rank")
domain_labels <- c("WIMD Overall","Income (22%)","Employment (20%)",
                   "Health (15%)","Education (14%)","Access (10%)",
                   "Housing (9%)","Community Safety (5%)","Physical Env (5%)")

map2_dfr(domain_cols, domain_labels, function(col, label) {
  t50 <- master %>% slice_min(.data[[col]], n = 50)
  tibble(domain         = label,
         rural_in_top50 = sum(t50$rural_urban == "Rural"),
         urban_in_top50 = sum(t50$rural_urban == "Urban"))
})
```




