---
title: "Domain weight gap assessment"
author: 'Anisha Bhattarai'
date: "2026-07-07"
output: html_document
---
<!-- This Script establish 19%/81% weight gap it works on every domain individually where rural score worse and income masking evidence-->
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)

```

## Step 1- Load dataset
```{r load_data}
master<- read_csv("../data/processed/master_dataset.csv") %>%
mutate(rural_urban = factor(rural_urban, levels = c("Rural", "Urban")))

rural <- master %>% filter(rural_urban == "Rural")
urban <- master %>% filter(rural_urban == "Urban")
N   <- nrow(master)
T20 <- floor(N * 0.20)
T10 <- floor(N * 0.10)

```

## Step 2- LSOAs % in worst 20%, each domain, rural vs urban
```{r domain_summary}
domain_cols <- c("income_rank","employment_rank","health_rank","education_rank",
                 "access_rank","housing_rank","community_safety_rank","physical_env_rank")
weights <- c(22,20,15,14,10,9,5,5)

map2_dfr(domain_cols, weights, function(col, w) {
  tibble(
    domain             = col,
    wimd_weight        = paste0(w, "%"),
    rural_pct_worst20  = round(100 * mean(rural[[col]] <= T20, na.rm=TRUE), 1),
    urban_pct_worst20  = round(100 * mean(urban[[col]] <= T20, na.rm=TRUE), 1),
    rural_median       = median(rural[[col]], na.rm=TRUE),
    urban_median       = median(urban[[col]], na.rm=TRUE),
    rural_worse        = median(rural[[col]], na.rm=TRUE) <
                         median(urban[[col]], na.rm=TRUE)
  )
}) %>% arrange(desc(rural_worse), desc(rural_pct_worst20))

```

## Step 3- Detail of Access domain
```{r access_domain}
master %>%
  group_by(rural_urban) %>%
  summarise(
    n            = n(),
    median_rank  = median(access_rank, na.rm=TRUE),
    pct_worst_20 = round(100 * mean(access_rank <= 383, na.rm=TRUE), 1),
    pct_worst_10 = round(100 * mean(access_rank <= 192, na.rm=TRUE), 1),
    .groups = "drop"
  )

```

## Step 4: Rural : worst 10% access vs rest, overall rank differ?
```{r access_invisible}
master %>%
  filter(rural_urban == "Rural") %>%
  mutate(access_group = if_else(access_rank <= 192,
                                "Worst 10% access", "Not worst 10%")) %>%
  group_by(access_group) %>%
  summarise(
    n                   = n(),
    median_overall_rank = median(wimd_2025_rank, na.rm=TRUE),
    median_income_rank  = median(income_rank,    na.rm=TRUE),
    pct_worst20_overall = round(100 * mean(wimd_2025_rank <= 383), 1),
    .groups = "drop"
  )

```

## Step 5- Detail of housing domain
```{r domain_housing}
master %>%
  group_by(rural_urban) %>%
  summarise(
    n            = n(),
    median_rank  = median(housing_rank, na.rm=TRUE),
    pct_worst_20 = round(100 * mean(housing_rank <= 383, na.rm=TRUE), 1),
    .groups = "drop"
  )
```

## Step 6- Rural : worst 10% access vs rest, overall rank differ?
```{r housing_rank_gap}

master %>%
  group_by(rural_urban) %>%
  summarise(
    n                  = n(),
    median_housing_gap = round(median(housing_rank_gap, na.rm=TRUE), 0),
    pct_negative       = round(100 * mean(housing_rank_gap < 0, na.rm=TRUE), 1),
    .groups = "drop"
  )

```

## Step 7- Housing deprivation in 4 local authorities
```{r housing_by_la}
master %>%
  filter(la_name %in% c("Powys","Ceredigion","Pembrokeshire","Swansea")) %>%
  group_by(la_name) %>%
  summarise(
    n              = n(),
    pct_worst20    = round(100 * mean(housing_rank <= T20, na.rm=TRUE), 1),
    median_housing = median(housing_rank, na.rm=TRUE),
    .groups = "drop"
  ) %>% arrange(pct_worst20)


```

## Step 8- Income, employment domain for comparision
```{r income_employment}
master %>%
  group_by(rural_urban) %>%
  summarise(
    n                      = n(),
    median_income_rank     = median(income_rank,     na.rm=TRUE),
    median_employment_rank = median(employment_rank, na.rm=TRUE),
    pct_worst20_income     = round(100 * mean(income_rank     <= T20, na.rm=TRUE), 1),
    pct_worst20_employment = round(100 * mean(employment_rank <= T20, na.rm=TRUE), 1),
    .groups = "drop"
  )

```

## Step 9- Income track the overall rank than access rank 
```{r income_masking}
cat("Within rural LSOAs — Spearman correlation of overall WIMD rank with:\n")
cat("  Income rank:", round(cor(rural$wimd_2025_rank, rural$income_rank, 
method = "spearman",                              use="complete.obs"), 3), "\n")
cat("  Access rank:", round(cor(rural$wimd_2025_rank, rural$access_rank, method = "spearman",                             use="complete.obs"), 3), "\n")

```

## Step 10- four domains
```{r remaining_domain}
map2_dfr(
  c("health_rank","education_rank","community_safety_rank","physical_env_rank"),
  c(15, 14, 5, 5),
  function(col, w) {
    tibble(
      domain            = col,
      wimd_weight       = paste0(w, "%"),
      rural_median      = median(rural[[col]], na.rm=TRUE),
      urban_median      = median(urban[[col]], na.rm=TRUE),
      rural_pct_worst20 = round(100 * mean(rural[[col]] <= T20, na.rm=TRUE), 1),
      urban_pct_worst20 = round(100 * mean(urban[[col]] <= T20, na.rm=TRUE), 1)
    )
  })

```

## Step 11- Weight- gap summary 
```{r weight_gap}
# 
tibble(
  domain      = domain_cols,
  wimd_weight = weights,
  rural_worse = c(FALSE,FALSE,FALSE,FALSE,TRUE,TRUE,FALSE,FALSE)
) %>%
  group_by(rural_worse) %>%
  summarise(
    group           = if_else(first(rural_worse), "Rural more deprived",
                              "Urban more deprived"),
    combined_weight = sum(wimd_weight),
    .groups = "drop"
  ) %>%
  select(group, combined_weight)

```

