---
title: "WIMD 2025 — Rural Deprivation in Wales"
author: "Anisha Bhattarai | Rural Health Compass | MSc Health Data Science"
date: "July 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
                      fig.width=9, fig.height=5, dpi=150)
library(tidyverse)

setwd("C:/Users/Bruger/Desktop/WIMD-Rural-deprivation-wales")

master <- read_csv("data/processed/master_dataset.csv") %>%
  mutate(
    rural_urban  = factor(rural_urban, levels=c("Rural","Urban")),
    drd_category = factor(drd_category,
      levels=c("LSOAs never ranked in top 50 most deprived",
               "LSOAs sometimes in the top 50 most deprived",
               "LSOAs often in the top 50 most deprived",
               "LSOAs in Deep-rooted deprivation"), ordered=TRUE))
rural <- master %>% filter(rural_urban=="Rural")
if(nrow(master)!=1917) stop("Check master_dataset.csv path")
cat("Ready:", nrow(master), "rows\n")

# Exact colours from HTML
RED  <- "#C0392B"
BLUE <- "#2471A3"
GRN  <- "#1a7a4a"
ORG  <- "#E67E22"
BG   <- "#FFFFFF"

# Theme matching HTML exactly
# HTML: grid:display:false on BAR axis, grid:color:GRID on VALUE axis
# border:display:false on both axes, no ticks
TH <- theme_minimal(base_size=11) +
  theme(
    plot.background   = element_rect(fill=BG, colour=NA),
    panel.background  = element_rect(fill=BG, colour=NA),
    plot.title        = element_text(face="bold", size=13, colour="#111111",
                                     margin=margin(b=3)),
    plot.subtitle     = element_text(size=9.5, colour="#52514E",
                                     margin=margin(b=8), lineheight=1.4),
    plot.caption      = element_text(size=7.5, colour="#898781",
                                     hjust=0, margin=margin(t=6)),
    plot.margin       = margin(14,18,10,14),
    axis.text         = element_text(size=9.5, colour="#444444"),
    axis.title        = element_text(size=9.5, colour="#444444"),
    axis.line         = element_blank(),
    axis.ticks        = element_blank(),
    legend.position   = "bottom",
    legend.background = element_rect(fill=BG, colour=NA),
    legend.key        = element_rect(fill=BG, colour=NA),
    legend.key.size   = unit(0.5,"cm"),
    legend.text       = element_text(size=9.5, colour="#52514E"),
    legend.title      = element_blank(),
    # HTML: value axis has grid, bar axis has no grid
    panel.grid.major  = element_line(colour="#EBEBEB", linewidth=0.4),
    panel.grid.minor  = element_blank()
  )
# No grid on categorical axis
TH_vcat <- TH + theme(panel.grid.major.x=element_blank()) # vertical bars
TH_hcat <- TH + theme(panel.grid.major.y=element_blank()) # horizontal bars

dir.create("outputs", showWarnings=FALSE)
sv <- function(p, f, h=5)
        ggsave(paste0("outputs/",f), p, width=9, height=h, dpi=150, bg=BG)
PCT <- function(x) paste0(x,"%")
SRC <- "Source: WIMD 2025 (Welsh Government); RUC 2021 (ONS). n=1,917 Welsh LSOAs."
```

## C1 — Which domains show the biggest rural-urban gap?
*Report 3.1 | Slide 4*

```{r c1}
# HTML: horizontal bar, indexAxis='y', barThickness=18
# single dataset, colour per bar (red if positive, blue if negative)
tibble(
  domain = c("Access (10%)","Housing (9%)","Physical Env (5%)",
             "Education (14%)","Employment (20%)","Overall WIMD",
             "Income (22%)","Health (15%)","Community Safety (5%)"),
  gap    = c(49.1,15.3,-15.1,-15.9,-16.4,-17.2,-18.6,-18.6,-19.2),
  col    = c(RED,RED,BLUE,BLUE,BLUE,BLUE,BLUE,BLUE,BLUE)
) %>%
  mutate(domain=fct_reorder(domain,gap)) %>%
  ggplot(aes(x=gap, y=domain, fill=col)) +
  geom_col(width=0.55, colour=BG, linewidth=0.3) +
  geom_vline(xintercept=0, linewidth=0.7, colour="#BBBBBB") +
  geom_text(aes(label=paste0(ifelse(gap>0,"+",""),gap,"%"),
                hjust=if_else(gap>0,-0.15,1.12)),
            size=3.2, fontface="bold", colour="#111111") +
  scale_fill_identity() +
  scale_x_continuous(limits=c(-26,60),
                     breaks=c(-20,-10,0,10,20,30,40,50),
                     labels=function(x) paste0(ifelse(x>0,"+",""),x,"%")) +
  labs(title    = "Which domains show the biggest rural-urban gap?",
       subtitle = "Red = rural more deprived  |  Blue = urban more deprived\nValue = percentage-point difference in areas in worst 20% of Wales",
       x=NULL, y=NULL, caption=SRC) +
  TH_hcat -> p1
p1; sv(p1,"c1_domain_gaps.png")
```

## C2 — WIMD weight mismatch ★
*Report 3.1 | Slide 5*

```{r c2}
# HTML: indexAxis='y', STACKED horizontal bar, ONE row showing 19% RED then 81% BLUE
# barThickness=40, max=100, y-axis ticks hidden
# This is a single horizontal bar — the most important structural chart
tibble(
  group   = "WIMD weight",
  rural   = 19,
  urban   = 81
) %>%
  pivot_longer(c(rural,urban), names_to="seg", values_to="pct") %>%
  mutate(seg=factor(seg, levels=c("rural","urban"),
                     labels=c("Rural-worse (19%): Access + Housing",
                              "Urban-worse (81%): Income + Employment + Health + Education + Safety + Env"))) %>%
  ggplot(aes(x=pct, y=group, fill=seg)) +
  geom_col(position=position_stack(reverse = TRUE), width=0.5, colour=BG, linewidth=0.5) +
  scale_fill_manual(values=c(
    "Rural-worse (19%): Access + Housing"                                                            =RED,
    "Urban-worse (81%): Income + Employment + Health + Education + Safety + Env"=BLUE)) +
  scale_x_continuous(limits=c(0,100), labels=PCT,
                     breaks=c(0,19,50,81,100)) +
  labs(title    = "WIMD gives 81% of its score to domains where urban Wales is more deprived",
       subtitle = "Rural Wales scores worst on domains worth only 19 marks out of 100.\nThis is why rural deprivation is hidden in the overall ranking.",
       x="Combined WIMD weight (%)", y=NULL,
       caption="Source: WIMD 2025 domain weights (Welsh Government); Stage 4 analysis.") +
  TH + theme(axis.text.y=element_blank(),
             panel.grid.major.y=element_blank()) -> p2
p2; sv(p2,"c2_weight_mismatch.png", h=3.2)
```

## C3 — Access deprivation is hidden
*Report 3.1 | Slide 6*

```{r c3}
# HTML: vertical bars, barThickness=70, max=100, RED and BLUE
tibble(
  group = c("Rural\n(561 areas)","Urban\n(1,356 areas)"),
  pct   = c(76.8,39.7),
  col   = c(RED,BLUE)
) %>%
  ggplot(aes(x=group, y=pct, fill=col)) +
  geom_col(width=0.44, colour=BG, linewidth=0.4) +
  geom_text(aes(label=PCT(pct)), colour=c(RED,BLUE),
            vjust=-0.4, size=13, fontface="bold") +
  scale_fill_identity() +
  scale_y_continuous(limits=c(0,92), labels=PCT, breaks=c(0,25,50,75)) +
  labs(title    = "76.8% of rural areas are more access-deprived than WIMD shows",
       subtitle = "The typical rural area is 680 rank positions worse on access than its overall WIMD rank suggests.",
       x=NULL, y="% of areas", caption=SRC) +
  TH_vcat -> p3
p3; sv(p3,"c3_access_hidden.png")
```

## C4 — Top 50 per domain: rural or urban? ★
*Report 3.1 | Slide 7*

```{r c4}
# HTML: stacked vertical bars, RED=rural, BLUE=urban, barThickness=24, max=50
tibble(
  domain = c("Access\n(10%)","Housing\n(9%)","Education\n(14%)",
             "Safety\n(5%)","Env\n(5%)","Employment\n(20%)",
             "Health\n(15%)","Income\n(22%)","Overall"),
  rural  = c(50,23,7,4,0,2,2,2,2),
  urban  = c(0,27,43,46,50,48,48,48,48)
) %>%
  mutate(domain=fct_reorder(domain,-rural)) %>%
  pivot_longer(c(rural,urban), names_to="group", values_to="n") %>%
  mutate(group=factor(group,levels=c("rural","urban"),
                       labels=c("Rural areas","Urban areas"))) %>%
  ggplot(aes(x=domain, y=n, fill=group)) +
  geom_col(position="stack", width=0.65, colour=BG, linewidth=0.4) +
  geom_text(aes(label=ifelse(n>=8,n,"")),
            position=position_stack(vjust=0.5),
            size=3.8, fontface="bold", colour="white") +
  scale_fill_manual(values=c("Rural areas"=RED,"Urban areas"=BLUE)) +
  scale_y_continuous(limits=c(0,52), breaks=c(0,25,50)) +
  labs(title    = "Of the 50 most deprived areas on each domain — rural or urban?",
       subtitle = "All 50 most access-deprived areas in Wales are Rural (10% weight).\nAll 50 most income-deprived are Urban (22% weight).",
       x=NULL, y="Number of areas (out of 50)", caption=SRC) +
  TH_vcat -> p4
p4; sv(p4,"c4_top50_domain.png")
```

## C5 — 20 years of tracking: rural Wales barely appears
*Report 3.2 | Slide 8*

```{r c5}
# HTML: stacked vertical bars, barThickness=70, max=100
# EXACT colours from HTML:
#   Never appeared  = #2471A3 (BLUE)
#   Sometimes       = #F39C12 (amber)
#   Often           = #E74C3C (red)
#   Deep-rooted     = #4d0000 (dark red)
# x-axis: grid:display:false | y-axis: grid shown
# TWO bars: Rural (left) and Urban (right)
drd_map <- c(
 "Never appeared" ="LSOAs never ranked in top 50 most deprived",
  "Sometimes" = "LSOAs sometimes in the top 50 most deprived",
  "Often" = "LSOAs often in the top 50 most deprived",
  "Deep-rooted" = "LSOAs in Deep-rooted deprivation"
)
p5 <- master %>%
  mutate(cat = fct_recode(drd_category, !!!drd_map) %>%
               fct_relevel("Never appeared","Sometimes","Often","Deep-rooted")) %>%
  count(rural_urban, cat) %>%
  group_by(rural_urban) %>%
  mutate(pct=round(100*n/sum(n),1)) %>%
  ggplot(aes(x=rural_urban, y=pct, fill=cat)) +
  geom_col(position="stack", width=0.4, colour=86, linewidth=0.4) +
  # Label only the "Never appeared" segment inside each bar
  geom_text(aes(label=ifelse(cat=="Never appeared",
                             paste0(n," areas\n(",pct,"%)"),"")),
            position=position_stack(vjust=0.5),
            size=3.5, fontface="bold", colour="white") +
  scale_fill_manual(
    values=c("Never appeared"="#2471A3",
             "Sometimes"     ="#F39C12",
             "Often"         ="#E74C3C",
             "Deep-rooted"   ="#4d0000")) +
  scale_y_continuous(labels=PCT, breaks=c(0,25,50,75,100)) +
  labs(title    = "20 years of tracking the most deprived areas — rural Wales barely appears",
       subtitle = "559 of 561 rural areas have NEVER appeared in the 50 most deprived since 2005.\nNot because they improved — access deprivation cannot push the overall score low enough.",
       x=NULL, y="% of areas",
       caption="Source: WIMD 2025 Deep-Rooted Deprivation (Welsh Government); RUC 2021 (ONS).") +
  TH_vcat + guides(fill=guide_legend(nrow=2))
p5; sv(p5,"c5_drd_tracking.png")
```

## C6 — Rural never-tracked and urban deep-rooted — opposite profiles
*Report 3.2 | Slide 9*

```{r c6}
# HTML: grouped vertical bars, BLUE=income, RED=access, barThickness=46, max=1400
tibble(
  group  = c("Rural areas\nnever tracked (559)","Urban areas\ndeep-rooted (20)"),
  income = c(1171,17),
  access = c(303,630)
) %>%
  pivot_longer(c(income,access), names_to="domain", values_to="rank") %>%
  mutate(domain=factor(domain,levels=c("income","access"),
                        labels=c("Income rank (22%)","Access rank (10%)"))) %>%
  ggplot(aes(x=group, y=rank, fill=domain)) +
  geom_col(position="dodge", width=0.54, colour=BG, linewidth=0.4) +
  geom_text(aes(label=paste0("Rank ",rank)),
            position=position_dodge(width=0.54),
            vjust=-0.4, size=3.8, fontface="bold", colour="#111111") +
  # Arrow to rank 17 — nearly invisible on 0-1400 scale
  annotate("text", x=1.73, y=230, label="Rank 17",
           size=3.5, fontface="bold", colour=BLUE) +
  annotate("segment", x=1.73, xend=1.73, y=170, yend=25,
           colour=BLUE, linewidth=0.8,
           arrow=arrow(length=unit(0.08,"cm"), type="closed")) +
  scale_fill_manual(values=c("Income rank (22%)"=BLUE,"Access rank (10%)"=RED)) +
  scale_y_continuous(limits=c(0,1400),
                     breaks=c(1,383,766,1150),
                     labels=c("1","383","766","1,150")) +
  labs(title    = "Rural never-tracked and urban deep-rooted — mirror-image profiles",
       subtitle = "Blue = income rank  |  Red = access rank  |  Lower = more deprived in Wales\nRural: income rank 1,171 (comfortable), access rank 303 (severely deprived).\nUrban deep-rooted: income rank 17 — most income-deprived in Wales.",
       x=NULL, y="Median rank (lower = more deprived)",
       caption=paste(SRC,"2025 ranks only.")) +
  TH_vcat -> p6
p6; sv(p6,"c6_drd_profiles.png")
```

## C7 — Income gets you visible. Access does not. ★
*Report 3.2 | Slide 10*

```{r c7}
# HTML: vertical bars, barThickness=56, max=100
# RED for income-related, GRN for access-only
tibble(
  group = c("Income deprived\nonly (33 areas)",
            "Both income\n& access (5 areas)",
            "Neither\n(221 areas)",
            "Access deprived\nonly (302 areas)"),
  pct   = c(87.9,80.0,3.2,1.3),
  col   = c(RED,RED,GRN,GRN)
) %>%
  mutate(group=fct_reorder(group,-pct)) %>%
  ggplot(aes(x=group, y=pct, fill=col)) +
  geom_col(width=0.6, colour=BG, linewidth=0.4) +
  geom_text(aes(label=PCT(pct)), colour=c(RED,RED,GRN,GRN),
            vjust=-0.4, size=7, fontface="bold") +
  scale_fill_identity() +
  scale_y_continuous(limits=c(0,104), labels=PCT, breaks=c(0,25,50,75,100)) +
  labs(title    = "Income deprivation gets a rural area onto the priority list. Access deprivation does not.",
       subtitle = "% of rural areas appearing in worst 20% of Wales overall, by type of deprivation.\nRed = income-deprived (visible)  |  Green = access-deprived only (hidden)",
       x="Type of deprivation  (rural areas only, n=561)",
       y="% appearing in worst 20% of Wales",
       caption=SRC) +
  TH_vcat -> p7
p7; sv(p7,"c7_mechanism.png")
```

## C8 — Four local authorities ★
*Report 3.4 | Slide 11*

```{r c8}
# HTML: grouped vertical bars, BLUE=Overall, RED=Access, ORG=Housing
# barThickness=20, max=80
tibble(
  la      = c("Powys","Ceredigion","Pembrokeshire","Swansea"),
  Overall = c(6.3,2.2,14.1,22.7),
  Access  = c(63.3,64.4,50.7,9.3),
  Housing = c(25.3,42.2,19.7,21.3)
) %>%
  pivot_longer(c(Overall,Access,Housing),
               names_to="measure", values_to="pct") %>%
  mutate(measure=factor(measure,levels=c("Overall","Access","Housing")),
         la=factor(la,levels=c("Powys","Ceredigion","Pembrokeshire","Swansea"))) %>%
  ggplot(aes(x=la, y=pct, fill=measure)) +
  geom_col(position="dodge", width=0.7, colour=BG, linewidth=0.3) +
  geom_text(aes(label=PCT(pct)),
            position=position_dodge(width=0.7),
            vjust=-0.4, size=3.2, fontface="bold", colour="#111111") +
  scale_fill_manual(values=c("Overall"=BLUE,"Access"=RED,"Housing"=ORG),
                    labels=c("Overall WIMD","Access (10%)","Housing (9%)")) +
  scale_y_continuous(limits=c(0,80), labels=PCT, breaks=c(0,20,40,60,80)) +
  labs(title    = "% of areas in most deprived 20% of Wales — four local authorities",
       subtitle = "Blue = Overall WIMD  |  Red = Access  |  Orange = Housing\nPowys and Ceredigion appear non-deprived overall (blue) but >60% worst 20% on access (red).",
       x=NULL, y="% of areas in worst 20% of Wales",
       caption=paste(SRC,"Worst 20% = rank ≤383 of 1,917.")) +
  TH_vcat -> p8
p8; sv(p8,"c8_case_studies.png")
```

## C9 — Inside Ceredigion: 802 rank positions apart
*Report 3.4 | Slide 12*

```{r c9}
# HTML: grouped vertical bars, BLUE=overall, RED=access, barThickness=52, max=1200
p9 <- tibble(
  area    = c("Aberystwyth\n(6 urban areas)","Rural Ceredigion\n(34 areas)"),
  Overall = c(732,872),
  Access  = c(936,134)
) %>%
  pivot_longer(c(Overall,Access), names_to="measure", values_to="rank") %>%
  mutate(measure=factor(measure,levels=c("Overall","Access"),
                         labels=c("Overall WIMD rank","Access rank"))) %>%
  ggplot(aes(x=area, y=rank, fill=measure)) +
  geom_col(position="dodge", width=0.52, colour=BG, linewidth=0.4) +
  geom_text(aes(label=paste0("Rank ",rank)),
            position=position_dodge(width=0.52),
            vjust=-0.4, size=3.8, fontface="bold", colour="#111111") +
  annotate("text", x=1.5, y=820, label="← 802 gap →",
           size=3.8, fontface="bold", colour=RED) +
  annotate("segment", x=1.73, xend=2.27, y=755, yend=755,
           colour=RED, linewidth=1,
           arrow=arrow(ends="both", length=unit(0.09,"cm"))) +
  scale_fill_manual(values=c("Overall WIMD rank"=BLUE,"Access rank"=RED)) +
  scale_y_continuous(limits=c(0,1200),
                     breaks=c(1,383,766,1150),
                     labels=c("1","383","766","1,150")) +
  labs(title    = "Inside Ceredigion: town and rural hinterland — 802 ranks apart on access",
       subtitle = "Aberystwyth access rank 936  vs  Rural Ceredigion access rank 134. Same local authority.",
       x=NULL, y="WIMD rank (1 = most deprived in Wales)",
       caption=SRC) +
  TH_vcat -> p9
p9; sv(p9,"c9_aberystwyth.png")
```

## C10 — Rural appears healthy but access tells a different story
*Report 3.3 | Slide 13*

```{r c10}
# HTML: grouped vertical bars, RED=Rural, BLUE=Urban, barThickness=60, max=70
tibble(
  measure = c("In BEST 20%\nfor Health","In WORST 20%\nfor Access"),
  Rural   = c(31.9,54.7),
  Urban   = c(18.1,5.6)
) %>%
  pivot_longer(c(Rural,Urban), names_to="group", values_to="pct") %>%
  mutate(group=factor(group,levels=c("Rural","Urban"))) %>%
  ggplot(aes(x=measure, y=pct, fill=group)) +
  geom_col(position="dodge", width=0.55, colour=BG, linewidth=0.4) +
  geom_text(aes(label=PCT(pct)),
            position=position_dodge(width=0.55),
            vjust=-0.4, size=5, fontface="bold",
            colour=rep(c(RED,BLUE),2)) +
  scale_fill_manual(values=c("Rural"=RED,"Urban"=BLUE)) +
  scale_y_continuous(limits=c(0,70), labels=PCT, breaks=c(0,20,40,60)) +
  labs(title    = "Rural Wales appears healthy — yet has the worst access to services",
       subtitle = "31.9% of rural areas are in the healthiest 20% of Wales.\nYet 54.7% are in the worst 20% on access — communities far from services may not register as health-deprived.",
       x=NULL, y="% of areas", caption=SRC) +
  TH_vcat -> p10
p10; sv(p10,"c10_health_lens.png")
```

## C11 — 190 communities invisible to policy
*Report Appendix | Slide 14*

```{r c11, fig.height=6}
# HTML: indexAxis='y' (horizontal bars), RED, barThickness=26, max=46
tibble(
  la = c("Powys","Carmarthenshire","Pembrokeshire","Ceredigion","Gwynedd",
          "Isle of Anglesey","Monmouthshire","Conwy","Denbighshire","Wrexham"),
  n  = c(39,26,25,21,21,11,11,9,8,5)
) %>%
  mutate(la=fct_reorder(la,n)) %>%
  ggplot(aes(x=n, y=la)) +
  geom_col(fill=RED, width=0.65, colour=BG, linewidth=0.3) +
  geom_text(aes(label=paste0(n," communities")),
            hjust=-0.08, size=3.8, fontface="bold", colour="#111111") +
  scale_x_continuous(limits=c(0,50), expand=c(0,0)) +
  labs(title    = "190 rural communities: worst 10% on access — invisible in WIMD overall rankings",
       subtitle = "0% appear in worst 20% overall. These communities will never trigger a WIMD-based policy response.",
       x="Number of communities", y=NULL,
       caption=paste("Invisible = access rank ≤192 AND overall rank >50.",SRC)) +
  TH_hcat -> p11
p11; sv(p11,"c11_invisible.png",h=6)
```

## C12 — 147 areas worst on both access and housing — still invisible
*Report 3.4 + Policy*

```{r c12, fig.height=6}
# HTML: indexAxis='y' (horizontal bars), ORG, barThickness=26, max=38
tibble(
  la  = c("Gwynedd","Carmarthenshire","Powys","Ceredigion",
           "Isle of Anglesey","Pembrokeshire","Denbighshire","Conwy"),
  n   = c(31,23,20,16,14,12,10,9),
  pct = c(55.4,42.6,29.0,47.1,42.4,22.2,33.3,47.4)
) %>%
  mutate(la=fct_reorder(la,n)) %>%
  ggplot(aes(x=n, y=la)) +
  geom_col(fill=ORG, width=0.65, colour=BG, linewidth=0.3) +
  geom_text(aes(label=paste0(n,"  (",pct,"% of rural ",la,")")),
            hjust=-0.04, size=3.4, colour="#111111") +
  scale_x_continuous(limits=c(0,64), expand=c(0,0)) +
  labs(title    = "147 rural areas worst 20% for BOTH access and housing — 0% visible overall",
       subtitle = "Compound deprivation on the two rural-specific domains. Not one appears in the overall worst 20%.",
       x="Number of areas", y=NULL,
       caption=paste("Both access rank ≤383 AND housing rank ≤383.",SRC)) +
  TH_hcat -> p12
p12; sv(p12,"c12_dual.png",h=6)
```
